Locally Conservative Discontinuous Petrov-Galerkin Finite Elements for Fluid Problems

### 2013-11-15 docs/Truman/LocalConservation/paper/paper.tex
Updates to local conservation paper

\documentclass[letterpaper]{article}
\usepackage[margin=1.0in]{geometry}
\usepackage{authblk}
\usepackage{amsmath, amssymb, amsthm, mathtools}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{color}
\usepackage{subcaption}

\def\btau{\boldsymbol\tau}
\def\bftau{\boldsymbol\tau}
\def\bsigma{\boldsymbol\sigma}
\def\bbeta{\boldsymbol\beta}
\def\blambda{\boldsymbol\lambda}

@@

\newcommand{\mb}[1]{\mathbf{#1}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\mc}[1]{\mathcal{#1}}
\newcommand{\norm}[1]{\left\| #1 \right\|}
\newcommand{\snorm}[1]{\left| #1 \right|}
\newcommand{\LRp}[1]{\left( #1 \right)}
\newcommand{\LRs}[1]{\left[ #1 \right]}
\newcommand{\LRa}[1]{\left\langle #1 \right\rangle}
\newcommand{\LRc}[1]{\left\{ #1 \right\}}
\newcommand{\tanbui}[2]{\textcolor{blue}{\sout{#1}} \textcolor{red}{#2}}
\newcommand{\Grad} {\ensuremath{\nabla}}
\newcommand{\Div} {\ensuremath{\nabla\cdot}}
\newcommand{\Nel} {\ensuremath{{N^\text{el}}}}
\newcommand{\jump}[1] {\ensuremath{\LRs{\![#1]\!}}}
\newcommand{\uh}{\widehat{u}}
\newcommand{\fnh}{\widehat{f}_n}
\renewcommand{\L}{L^2\LRp{\Omega}}
\newcommand{\pO}{\partial\Omega}
\newcommand{\Gh}{\Gamma_h}
\newcommand{\Gm}{\Gamma_{-}}
\newcommand{\Gp}{\Gamma_{+}}
\newcommand{\Go}{\Gamma_0}
\newcommand{\Oh}{\Omega_h}
\newcommand{\ptl}{{\partial}}
\newcommand{\bfsig}{\mbox{\boldmath $\sigma$}}
\newcommand{\bfv}{\boldsymbol v}
\newcommand{\bfu}{\boldsymbol u}
\newcommand{\bfpsi}{\boldsymbol\psi}
\newcommand{\bflambda}{\boldsymbol\lambda}
\newcommand{\bfn}{\mbox{\boldmath $n$}}
\newcommand{\bfH}{\mbox{\boldmath $H$}}
\newcommand{\HdivK}{\bfH(\text{div},K)}
\newcommand{\HOneK}{H^{1}(K)}
\newcommand{\HOneOmegah}{H^{-1}(\Omega_h)}
\newcommand{\HdivOmegah}{\bfH(\text{div},\Omega_h)}
\newcommand{\vdeltau}{v_{\delta\bs u_h}}
\newcommand{\taudeltau}{\btau_{\delta\bs u_h}}

@@

\def\arr#1#2#3#4{\left[
\begin{array}{cc}
#1 & #2\\
#3 & #4\\
\end{array}
\right]}
\def\vecttwo#1#2{\left[
\begin{array}{c}
#1\\
#2\\
\end{array}
\right]}
\def\vectthree#1#2#3{\left[
\begin{array}{c}
#1\\
#2\\
#3\\
\end{array}
\right]}
\def\vectfour#1#2#3#4{\left[
\begin{array}{c}
#1\\
#2\\
#3\\
#4\\
\end{array}
\right]}

\renewcommand{\arraystretch}{2.0}

@@

\subsection{Derivation}
Now that we have briefly outlined the abstract DPG method, let us apply it to
the convection-diffusion equation. The strong form of the steady
convection-diffusion problem with homogeneous Dirichlet boundary conditions reads
\[
\left\{
\begin{array}[c]{rrl}
\div(\bs\beta u)-\epsilon\del u & =f & \text{in }\Omega\\
u & =0 & \text{on }\Gamma\,,
\end{array}
\right.
\]
where $u$ is the property of interest, $\bs\beta$ is the convection vector,
and $f$ is the source term. Nonhomogeneous Dirichlet and Neumann boundary
conditions are straightforward but would add technicality to the following
discussion. Let us write this as an equivalent system of first
order equations:
\begin{align*}
\div(\bs\beta u-\bs\sigma)&=f\\
\frac{1}{\epsilon}\bs\sigma-\grad u&=\bs0\,.
\end{align*}
If we then multiply the top equation by some scalar test function $v$ and the
bottom equation by some vector-valued test function $\btau$, we can integrate by
parts over each element $K$:
\begin{equation}
\label{eq:preultraweak}
\begin{aligned}
-(\bbeta u-\bsigma,\nabla v)_K+((\bbeta
u-\bsigma)\cdot\mathbf{n},v)_{\partial K}&=(f,v)_K\\
\frac{1}{\epsilon}(\bsigma,\btau)_K+(u,\nabla\cdot\btau)_K
-(u,\tau_n)_{\partial K}&=0\,.
\end{aligned}
\end{equation}
The discontinuous Petrov-Galerkin method refers to the fact that we are using
discontinuous optimal test functions that come from a space differing from the
trial space. It does not specify our choice of trial space. Nevertheless, many
considerations of DPG in the literature (convection-diffusion \cite{DPG6},
linear elasticity \cite{BramwellDemkowiczGopalakrishnanQiu11}, linear
acoustics \cite{DemkowiczGopalakrishnanMugaZitelli12}, Stokes
\cite{DPGStokes}) associate DPG with the so-called ``ultra-weak formulation.''
We will follow the same derivation for the convection-diffusion equation, but
we emphasize that other formulations are available (in particular, the
Primal DPG\cite{PrimalDPG} method presents an alternative with
continuous trial functions). Thus, we seek field variables $u\in L^2(K)$ and
$\bsigma\in\mb{L^2}(K)$. Mathematically, this leaves their traces on element
boundaries undefined, and in a manner similar to the hybridized discontinuous
Galerkin method, we define new unknowns for trace $\hat u$ and flux $\hat t$.
Applying these definitions to \eqnref{preultraweak} and adding the two
equations together, we arrive at our desired variational problem.

Find
$\bs u:=(u,\bsigma,\hat u,\hat t)
\in\bs U:=L^2(\Omega_h)\times \bs L^2(\Omega_h)\times H^{1/2}(\Gamma_h)\times H^{-1/2}(\Gamma_h)$
such that
\begin{align}
\label{eq:variationalFormulation}
\underbrace{-(\bbeta u-\bsigma,\nabla v)_K+(\hat t,v)_{\partial K}
+ \frac{1}{\epsilon}(\bsigma,\btau)_K
+(u,\nabla\cdot\btau)_K
-(\hat u,\tau_n)_{\partial K}}_{b(\mathbf{u}, \mathbf{v})}
&=\underbrace{\:(f,v)_K\genfrac{}{}{0pt}{}{}{}}_{l(\mathbf{v})} &\text{in }\Omega \\
\hat u&=0 &\text{on }\Gamma
\end{align}
for all $\bs v:=(v,\btau)\in
\bs V:=H^1(\Omega_h)\times\bfH(\text{div},\Omega_h)$.

Let $\bs U_h:=U_h\times\bs S_h\times\hat U_h\times\hat F_h\subset L^2(\Omega_h)\times\bs
L^2(\Omega_h)\times H^{\frac{1}{2}}(\Gamma_h)\times H^{-\frac{1}{2}}(\Gamma_h)$
be a finite-dimensional subspace, and let $\bs u_h:=(u_h.\bsigma_h,\hat
u_h\hat t_h)\in\bs U_h$ be a group variable. The element conservative DPG scheme is
derived from the Lagrangian:
\begin{equation}
L(\bs u_h,\lambda_k)=\frac{1}{2}\norm{R_V^{-1}(b(\bs
u_h,\cdot)-(f,\cdot))}^2_{\bs V}-\sum_K\lambda_K(b(\bs u_h,(1_K,\bs0))-l((1_K,\bs0)))\,,
\label{eq:lagrangian}
\end{equation}
where $(1_K,\bs0)$ is the test function in which $v=1$ on element $K$ and 0 elsewhere and $\btau=\bs0$ everywhere.

@@

Then, putting \eqref{eq:conservativeSystem} into more concrete terms for
convection-diffusion, we get:
\begin{equation}
\left\{
\begin{array}[c]{rll}
-(\bbeta u-\bsigma,\nabla \vdeltau)+\langle\hat t,\vdeltau\rangle
+ \frac{1}{\epsilon}(\bsigma,\taudeltau)
+(u,\nabla\cdot\taudeltau)
-\langle\hat u,\taudeltau\cdot\bs n\rangle\\
-\sum_K\lambda_K (\delta\hat t,(1_K,\bs0))
&=(f,\vdeltau) & \forall\delta\bs u_h\in\bs U_h\\
\langle\hat t,(1_K,\bs0)\rangle &=(f,1_K) & \forall K\,.
\end{array}
\right.
\label{eq:conservativeConfusionSystem}
\end{equation}

@@

\section{Robust test norms}
The optimal test functions are determined by solving local problems determined
by the choice of test norm. There are several options to consider.
The graph norm \cite{DPGOverview} is one of the most natural norms to consider
as it is derived directly from the adjoint of the problem supplemented with
(possibly scaled) $L^2$ field terms to upgrade it from a semi-norm.
Chan \etal\cite{ChanHeuerThanhDemkowicz2012} derived a more robust alternative
norm for convection diffusion (dubbed the robust norm). We recently
developed a modification of the robust norm that appears to produce better
results in the presence of singularities. In this section we illustrate the
singularity issues with the Laplace equation, offer possible explanations for
the phenomena observed, and propose a modification of the robust test norm,
which we demonstrate eliminates the issues observed in numerical experiments.

\subsection{A model problem}
\seclab{sec:confusionPlate}
We begin by first examining a different problem than convection-diffusion.
We look at admissible solutions for the homogeneous Laplace's equation
(convection diffusion with $\bbeta=\bs 0$) over
the $y > 0$ half-plane under boundary conditions
\begin{align*}
u &= 0 \text{ on } x > 0\\
\pd{u}{n} &= 0 \text{ on } x < 0.
\end{align*}
Let us consider the 2D case - a simple separation of variables argument in
polar coordinates shows that the solution is of the form
\[
u(r,\theta) = \sum_{n=0}^\infty R_n(r) \sin(\lambda_n \theta),
\]
where $\lambda_n = n + \frac{1}{2}$, and $R_n(r) = C_{1,n}x^{\lambda_n} +
C_{2,n}x^{-\lambda_n}$.  By requiring $u(0,\theta) < \infty$, we have $R_n(r)
= C_n x^{\lambda_n}$.  We have now that solutions to this problem include $u$
of the form
\[
u = \sum_{n=0} x^{n+\frac{1}{2}} \sin\LRp{\LRp{n+\frac{1}{2}}\theta}.
\]
Note that even in the lowest-order term, the gradient of $u$ displays a
singularity at $r = 0$. It is well known that, for smooth boundary data,
solutions to Laplace’s equation can be decomposed into the linear combination
of smooth and singular contributions; the above analysis implies that, when
boundary conditions change from Dirichlet to Neumann on the half-plane, the
Laplace’s equation will always develop a singularity in the stresses.

@@

We note that the appearance of this non-physical singularity in $u$ is allowed
under the theory underlying the robust test norm; the error in the $\L$-norm
of the solution is guaranteed to be robustly bounded; however, the $\L$ norm
does allow for the presence of weak singularities (singularities of order
$x^{-\frac{1}{2}}$).  Apart from the oscillation of $u$ at the singular point,
the solution is well-behaved, and the stress $\bsigma = \epsilon \Grad u$ is
very well represented, as indicated in Figure~\ref{fig:plateStresses}.

@@

\subsection{A modification of the robust test norm}

@@

While not rigorously understood, we believe the issues related to the
appearance of non-physical singularities to be related to the uncoupled nature
of the test norm.  Previous example problems exhibited boundary layers and
sharp gradients in the stress $\bsigma$, but not singularities, which
contribute significantly more error.  We expect that the oscillations observed
in $u$ are a sort of \textit{pollution error}, where error in $u$ is tied to
error in $\bsigma$.  If we consider the ultra-weak variational formulation for
convection-diffusion
\[
\LRp{u,\div \btau - \bbeta\cdot \grad v}_{\L} + \LRp{\bsigma,\frac{1}{\epsilon}
\btau + \grad v}_{\L} + \ldots,
\]
we can see that it is a combination of test functions that corresponds to both
$u$ and $\bsigma$.  Referring to \cite{ChanHeuerThanhDemkowicz2012}, we note
that, by choosing $\btau$
and $v$ such that they satisfy the adjoint equation with forcing terms $u$ and
$\bsigma$, we recover the best $L^2$ approximation.  In other words, achieving
optimality in the $L^2$ norm requires coupling between $v$ and $\btau$, which
is achieved under the graph norm, but not the robust norm derived in the
previous section.
If coupling of the test terms delivers optimality in $u$ and $\bsigma$
independently, we expect that decoupling
$v$ and $\btau$ from each other in a test norm will have the
effect of coupling error in $\bsigma$ to error in $u$, which would explain the
spurious oscillations in $u$ in the presence of singularities in $\bsigma$.
Similar results have been observed in the Stokes equations, where error in $u$
is coupled to the behavior of the pressure variable \cite{FLD:FLD582}.

@@

Figure~\ref{fig:newNorm} shows the solution for $\epsilon = .01$, where the
diffusion scale is both underresolved and resolved by $h$-adaptivity.  In both
cases, there are no oscillations near the plate tip --
Figure~\ref{fig:newNormZoom} shows a zoomed image of the solution $u$ at the
point $(.5,0)$.  The stress is resolved similarly to the previous case;
however, the solution $u$ does not display spurious oscillations in either the
underresolved or resolved cases.  Figure~\ref{fig:newNormSmallEps} displays
the same quantities, but for $\epsilon = 10^{-4}$, in order to demonstrate
that the new test norm removes spurious oscillations in $u$ (in the presence
of singularities in $\bsigma$) independently of $\epsilon$.

@@

\subsection{Adaptation for a Locally Conservative Formulation}
With this choice of test norm, our local problem now becomes:

@@

\section{Numerical Experiments}
In \ref{sec:problemDescriptions} we define each numerical experiment, and in
\ref{sec:problemAnalysis} we discuss the solution properties in general. We
solve with second order field variables and flux ($u$, $\bsigma$, and $\hat
t$), third order traces ($\hat t$), and fifth order test functions ($v$ and
$\btau$).

@@

\subsection{Description of Problems}\label{sec:problemDescriptions}
Unless otherwise noted, the problem domain is $\Omega=[0,1]^2$ and $f=0$. Also
note that except for the discontinuous source problem, for all of the
pseudo-color plots, blue corresponds to $0$ and red to $1$ with a linear
scaling in between. Also, all plots are of the field variable $u$.

@@

>>>>

\documentclass[letterpaper]{article}
\usepackage[margin=1.0in]{geometry}
\usepackage{authblk}
\usepackage{amsmath, amssymb, amsthm, mathtools}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{color}
\usepackage{subcaption}
\usepackage{morefloats}
\usepackage{pgf,tikz}
\usetikzlibrary{arrows}

\def\btau{\boldsymbol\tau}
\def\bftau{\boldsymbol\tau}
\def\bfsigma{\boldsymbol\sigma}
\def\bbeta{\boldsymbol\beta}
\def\blambda{\boldsymbol\lambda}

@@

\newcommand{\mb}[1]{\mathbf{#1}}
\newcommand{\mbb}[1]{\mathbb{#1}}
\newcommand{\mc}[1]{\mathcal{#1}}
\newcommand{\norm}[1]{\left\| #1 \right\|}
\newcommand{\snorm}[1]{\left| #1 \right|}
\newcommand{\LRp}[1]{\left( #1 \right)}
\newcommand{\LRs}[1]{\left[ #1 \right]}
\newcommand{\LRa}[1]{\left\langle #1 \right\rangle}
\newcommand{\LRc}[1]{\left\{ #1 \right\}}
\newcommand{\tanbui}[2]{\textcolor{blue}{\sout{#1}} \textcolor{red}{#2}}
\newcommand{\Grad} {\ensuremath{\nabla}}
\newcommand{\Div} {\ensuremath{\nabla\cdot}}
\newcommand{\Nel} {\ensuremath{{N^\text{el}}}}
\newcommand{\jump}[1] {\ensuremath{\LRs{\![#1]\!}}}
\newcommand{\uh}{\widehat{u}}
\newcommand{\fnh}{\widehat{f}_n}
\renewcommand{\L}{L^2\LRp{\Omega}}
\newcommand{\pO}{\partial\Omega}
\newcommand{\Gh}{\Gamma_h}
\newcommand{\Gm}{\Gamma_{-}}
\newcommand{\Gp}{\Gamma_{+}}
\newcommand{\Go}{\Gamma_0}
\newcommand{\Oh}{\Omega_h}
\newcommand{\ptl}{{\partial}}
\newcommand{\bfsig}{\mbox{\boldmath $\sigma$}}
\newcommand{\bfv}{\boldsymbol v}
\newcommand{\bfu}{\boldsymbol u}
\newcommand{\bff}{\boldsymbol f}
\newcommand{\bft}{\boldsymbol t}
\newcommand{\bfpsi}{\boldsymbol\psi}
\newcommand{\bflambda}{\boldsymbol\lambda}
\newcommand{\bfn}{\mbox{\boldmath $n$}}
\newcommand{\bfH}{\mbox{\boldmath $H$}}
\newcommand{\HdivK}{\bfH(\text{div},K)}
\newcommand{\HOneK}{H^{1}(K)}
\newcommand{\HOneOmegah}{H^{-1}(\Omega_h)}
\newcommand{\HdivOmegah}{\bfH(\text{div},\Omega_h)}
\newcommand{\vdeltau}{v_{\delta\bs u_h}}
\newcommand{\taudeltau}{\btau_{\delta\bs u_h}}

@@

\def\arr#1#2#3#4{\left[
\begin{array}{cc}
#1 & #2\\
#3 & #4\\
\end{array}
\right]}
\def\vecttwo#1#2{\left(
\begin{array}{c}
#1\\
#2\\
\end{array}
\right)}
\def\vectthree#1#2#3{\left[
\begin{array}{c}
#1\\
#2\\
#3\\
\end{array}
\right]}
\def\vectfour#1#2#3#4{\left[
\begin{array}{c}
#1\\
#2\\
#3\\
#4\\
\end{array}
\right]}

% \renewcommand{\arraystretch}{2.0}

@@

\subsection{Derivation}
Now that we have briefly outlined the abstract DPG method, let us apply it to
the convection-diffusion equation. The strong form of the steady
convection-diffusion problem with homogeneous Dirichlet boundary conditions reads
\[
\left\{
\begin{array}[c]{rrl}
\div(\bs\beta u)-\epsilon\del u & =f & \text{in }\Omega\\
u & =0 & \text{on }\Gamma\,,
\end{array}
\right.
\]
where $u$ is the property of interest, $\bs\beta$ is the convection vector,
and $f$ is the source term. Nonhomogeneous Dirichlet and Neumann boundary
conditions are straightforward but would add technicality to the following
discussion. Let us write this as an equivalent system of first
order equations:
\begin{align*}
\div(\bs\beta u-\bs\sigma)&=f\\
\frac{1}{\epsilon}\bs\sigma-\grad u&=\bs0\,.
\end{align*}
If we then multiply the top equation by some scalar test function $v$ and the
bottom equation by some vector-valued test function $\btau$, we can integrate by
parts over each element $K$:
\begin{equation}
\label{eq:preultraweak}
\begin{aligned}
-(\bbeta u-\bfsigma,\nabla v)_K+((\bbeta
u-\bfsigma)\cdot\mathbf{n},v)_{\partial K}&=(f,v)_K\\
\frac{1}{\epsilon}(\bfsigma,\btau)_K+(u,\nabla\cdot\btau)_K
-(u,\tau_n)_{\partial K}&=0\,.
\end{aligned}
\end{equation}
The discontinuous Petrov-Galerkin method refers to the fact that we are using
discontinuous optimal test functions that come from a space differing from the
trial space. It does not specify our choice of trial space. Nevertheless, many
considerations of DPG in the literature (convection-diffusion \cite{DPG6},
linear elasticity \cite{BramwellDemkowiczGopalakrishnanQiu11}, linear
acoustics \cite{DemkowiczGopalakrishnanMugaZitelli12}, Stokes
\cite{DPGStokes}) associate DPG with the so-called ``ultra-weak formulation.''
We will follow the same derivation for the convection-diffusion equation, but
we emphasize that other formulations are available (in particular, the
Primal DPG\cite{PrimalDPG} method presents an alternative with
continuous trial functions). Thus, we seek field variables $u\in L^2(K)$ and
$\bfsigma\in\mb{L^2}(K)$. Mathematically, this leaves their traces on element
boundaries undefined, and in a manner similar to the hybridized discontinuous
Galerkin method, we define new unknowns for trace $\hat u$ and flux $\hat t$.
Applying these definitions to \eqnref{preultraweak} and adding the two
equations together, we arrive at our desired variational problem.

Find
$\bs u:=(u,\bfsigma,\hat u,\hat t)
\in\bs U:=L^2(\Omega_h)\times \bs L^2(\Omega_h)\times H^{1/2}(\Gamma_h)\times H^{-1/2}(\Gamma_h)$
such that
\begin{align}
\label{eq:variationalFormulation}
\underbrace{-(\bbeta u-\bfsigma,\nabla v)_K+(\hat t,v)_{\partial K}
+ \frac{1}{\epsilon}(\bfsigma,\btau)_K
+(u,\nabla\cdot\btau)_K
-(\hat u,\tau_n)_{\partial K}}_{b(\mathbf{u}, \mathbf{v})}
&=\underbrace{\:(f,v)_K\genfrac{}{}{0pt}{}{}{}}_{l(\mathbf{v})} &\text{in }\Omega \\
\hat u&=0 &\text{on }\Gamma
\end{align}
for all $\bs v:=(v,\btau)\in
\bs V:=H^1(\Omega_h)\times\bfH(\text{div},\Omega_h)$.

Let $\bs U_h:=U_h\times\bs S_h\times\hat U_h\times\hat F_h\subset L^2(\Omega_h)\times\bs
L^2(\Omega_h)\times H^{\frac{1}{2}}(\Gamma_h)\times H^{-\frac{1}{2}}(\Gamma_h)$
be a finite-dimensional subspace, and let $\bs u_h:=(u_h.\bfsigma_h,\hat
u_h\hat t_h)\in\bs U_h$ be a group variable. The element conservative DPG scheme is
derived from the Lagrangian:
\begin{equation}
L(\bs u_h,\lambda_k)=\frac{1}{2}\norm{R_V^{-1}(b(\bs
u_h,\cdot)-(f,\cdot))}^2_{\bs V}-\sum_K\lambda_K(b(\bs u_h,(1_K,\bs0))-l((1_K,\bs0)))\,,
\label{eq:lagrangian}
\end{equation}
where $(1_K,\bs0)$ is the test function in which $v=1$ on element $K$ and 0 elsewhere and $\btau=\bs0$ everywhere.

@@

Then, putting \eqref{eq:conservativeSystem} into more concrete terms for
convection-diffusion, we get:
\begin{equation}
\left\{
\begin{array}[c]{rll}
-(\bbeta u-\bfsigma,\nabla \vdeltau)+\langle\hat t,\vdeltau\rangle
+ \frac{1}{\epsilon}(\bfsigma,\taudeltau)
+(u,\nabla\cdot\taudeltau)
-\langle\hat u,\taudeltau\cdot\bs n\rangle\\
-\sum_K\lambda_K (\delta\hat t,(1_K,\bs0))
&=(f,\vdeltau) & \forall\delta\bs u_h\in\bs U_h\\
\langle\hat t,(1_K,\bs0)\rangle &=(f,1_K) & \forall K\,.
\end{array}
\right.
\label{eq:conservativeConfusionSystem}
\end{equation}

@@

\subsection{Robust test norms}
The optimal test functions are determined by solving local problems determined
by the choice of test norm. There are several options to consider.
The graph norm \cite{DPGOverview} is one of the most natural norms to consider
as it is derived directly from the adjoint of the problem supplemented with
(possibly scaled) $L^2$ field terms to upgrade it from a semi-norm.
Chan \etal\cite{ChanHeuerThanhDemkowicz2012} derived a more robust alternative
norm for convection diffusion (dubbed the robust norm). We recently
developed a modification of the robust norm that appears to produce better
results in the presence of singularities. In this section we illustrate the
singularity issues with the Laplace equation, offer possible explanations for
the phenomena observed, and propose a modification of the robust test norm,
which we demonstrate eliminates the issues observed in numerical experiments.

\subsubsection{A model problem}
\seclab{sec:confusionPlate}
We begin by first examining a different problem than convection-diffusion.
We look at admissible solutions for the homogeneous Laplace's equation
(convection diffusion with $\bbeta=\bs 0$) over
the $y > 0$ half-plane under boundary conditions
\begin{align*}
u &= 0 \text{ on } x > 0\\
\pd{u}{n} &= 0 \text{ on } x < 0.
\end{align*}
Let us consider the 2D case - a simple separation of variables argument in
polar coordinates shows that the solution is of the form
\[
u(r,\theta) = \sum_{n=0}^\infty R_n(r) \sin(\lambda_n \theta),
\]
where $\lambda_n = n + \frac{1}{2}$, and $R_n(r) = C_{1,n}x^{\lambda_n} +
C_{2,n}x^{-\lambda_n}$.  By requiring $u(0,\theta) < \infty$, we have $R_n(r)
= C_n x^{\lambda_n}$.  We have now that solutions to this problem include $u$
of the form
\[
u = \sum_{n=0} x^{n+\frac{1}{2}} \sin\LRp{\LRp{n+\frac{1}{2}}\theta}.
\]
Note that even in the lowest-order term, the gradient of $u$ displays a
singularity at $r = 0$. It is well known that, for smooth boundary data,
solutions to Laplace’s equation can be decomposed into the linear combination
of smooth and singular contributions; the above analysis implies that, when
boundary conditions change from Dirichlet to Neumann on the half-plane, the
Laplace’s equation will always develop a singularity in the stresses.

@@

We note that the appearance of this non-physical singularity in $u$ is allowed
under the theory underlying the robust test norm; the error in the $\L$-norm
of the solution is guaranteed to be robustly bounded; however, the $\L$ norm
does allow for the presence of weak singularities (singularities of order
$x^{-\frac{1}{2}}$).  Apart from the oscillation of $u$ at the singular point,
the solution is well-behaved, and the stress $\bfsigma = \epsilon \Grad u$ is
very well represented, as indicated in Figure~\ref{fig:plateStresses}.

@@

\subsubsection{A modification of the robust test norm}

@@

While not rigorously understood, we believe the issues related to the
appearance of non-physical singularities to be related to the uncoupled nature
of the test norm.  Previous example problems exhibited boundary layers and
sharp gradients in the stress $\bfsigma$, but not singularities, which
contribute significantly more error.  We expect that the oscillations observed
in $u$ are a sort of \textit{pollution error}, where error in $u$ is tied to
error in $\bfsigma$.  If we consider the ultra-weak variational formulation for
convection-diffusion
\[
\LRp{u,\div \btau - \bbeta\cdot \grad v}_{\L} + \LRp{\bfsigma,\frac{1}{\epsilon}
\btau + \grad v}_{\L} + \ldots,
\]
we can see that it is a combination of test functions that corresponds to both
$u$ and $\bfsigma$.  Referring to \cite{ChanHeuerThanhDemkowicz2012}, we note
that, by choosing $\btau$
and $v$ such that they satisfy the adjoint equation with forcing terms $u$ and
$\bfsigma$, we recover the best $L^2$ approximation.  In other words, achieving
optimality in the $L^2$ norm requires coupling between $v$ and $\btau$, which
is achieved under the graph norm, but not the robust norm derived in the
previous section.
If coupling of the test terms delivers optimality in $u$ and $\bfsigma$
independently, we expect that decoupling
$v$ and $\btau$ from each other in a test norm will have the
effect of coupling error in $\bfsigma$ to error in $u$, which would explain the
spurious oscillations in $u$ in the presence of singularities in $\bfsigma$.
Similar results have been observed in the Stokes equations, where error in $u$
is coupled to the behavior of the pressure variable \cite{FLD:FLD582}.

@@

Figure~\ref{fig:newNorm} shows the solution for $\epsilon = .01$, where the
diffusion scale is both underresolved and resolved by $h$-adaptivity.  In both
cases, there are no oscillations near the plate tip --
Figure~\ref{fig:newNormZoom} shows a zoomed image of the solution $u$ at the
point $(.5,0)$.  The stress is resolved similarly to the previous case;
however, the solution $u$ does not display spurious oscillations in either the
underresolved or resolved cases.  Figure~\ref{fig:newNormSmallEps} displays
the same quantities, but for $\epsilon = 10^{-4}$, in order to demonstrate
that the new test norm removes spurious oscillations in $u$ (in the presence
of singularities in $\bfsigma$) independently of $\epsilon$.

@@

\subsubsection{Adaptation for a Locally Conservative Formulation}
With this choice of test norm, our local problem now becomes:

@@

\section{Application to Other Fluid Model Problems}
Extension of these ideas to other fluid flow problems is relatively trivial.
For the following problems, we just use the graph norm for the local problems.

\subsection{Inviscid Burgers' Equation}
The inviscid Burger's equation is
\[
\frac{\partial u}{\partial t}+u\frac{\partial u}{\partial x}=f\,.
\]
Define the space-time gradient: $\Grad_{xt}=\LRp{\frac{\partial}{\partial
x},\frac{\partial}{\partial t}}^T$. We can now rewrite this as
\[
\vecttwo{u}{1}\cdot\Grad_{xt}u=0\,.
\]
Multiplying by a test function $v$, and integrating by parts:
\[
-\LRp{\vecttwo{u}{1}u,\Grad_{xt}v}+\LRa{\hat t,v}=\LRp{f,v}\,,
\]
where $\hat t$ is the trace of $\vecttwo{u^2}{u}\cdot\mathbf{n}_{xt}$ on
element boundaries, and $\mathbf{n}_{xt}$ is the space-time normal vector.
As in convection-diffusion, local conservation implies that
$\int_{\partial K}\hat t=\int_Kf$ for all elements, $K$.

In order to solve this nonlinear problem, we linearize and do a simple Newton iteration
until the solution converges. The results follow in Section \ref{sec:inviscidBurgers}.

\subsection{Stokes Flow}
We start with the VGP (velocity, gradient pressure) Stokes formulation:
\begin{align*}
   \mu\Delta \bfu+\Grad p&=\bff\\
   \Div\bfu&=0\,,
\end{align*}
where $\bfu$ is the velocity vector field.
% For the following discussion, we let $\mu=1$, though this is by no means
% essential.
As a first order system of equations, this is
\begin{align*}
   \frac{1}{\mu}\bfsigma-\Grad\bfu&=0\\
   \Div\bfsigma+\Grad p&=\bff\\
   \Div\bfu&=0\,,
\end{align*}
where $\bfsigma$ is a tensor valued stress field.
Multiplying by test functions $\bftau$ (tensor valued), $\bfv$ (vector
valued), and $q$ (scalar valued), and integrating by parts:
\begin{align*}
   \LRp{\frac{1}{\mu}\bfsigma,\bftau}+\LRp{\bfu,\Div\bftau}-\LRa{\hat
   \bfu,\bftau\cdot\mathbf{n}}&=0\\
   -\LRp{\bfsigma,\Grad\bfv}-\LRp{p,\Div\bfv}+\LRa{\hat \bft,\bfv}&=\LRp{\bff,\bfv}\\
   -\LRp{\bfu,\Grad q}+\LRa{\hat\bfu\cdot\mathbf{n},q}&=0\,,
\end{align*}
where $\hat\bfu$ is the trace of $\bfu$, and $\hat\bft$ is the trace of
$\LRp{\bfsigma+p\mathbf{I}}\cdot\mathbf{n}$. The solve for $p$ is only unique
up to a constant, so we also impose a zero mean condition, $\int_\Omega p=0$.
Local conservation for Stokes flow means that over each element,
$\int_K\hat\bfu\cdot\mathbf{n}=0$. Results follow in Sections
\ref{sec:stokesCylinder} and \ref{sec:stokesStep}.

\section{Numerical Experiments}
In \ref{sec:problemDescriptions} we define each numerical experiment, and in
\ref{sec:problemAnalysis} we discuss the solution properties in general. We
solve with second order field variables and flux ($u$, $\bfsigma$, and $\hat
t$), third order traces ($\hat t$), and fifth order test functions ($v$ and
$\btau$).

@@

\subsection{Description of Problems}\label{sec:problemDescriptions}
Unless otherwise noted, the problem domain is $\Omega=[0,1]^2$ and $f=0$. Also
note that unless otherwise noted, for all of the
pseudo-color plots, blue corresponds to $0$ and red to $1$ with a linear
scaling in between. Also, all convection-diffusion plots are of the field variable $u$.
Inviscid Burgers' and Stokes results will be dealt with individually.

@@

\subsubsection{Inviscid Burgers' Equation}\label{sec:inviscidBurgers}

\subsubsection{Stokes Flow Around a Cylinder}\label{sec:stokesCylinder}
This is a common problem used to stress-test local conservation properties of
least squares finite element methods. Since DPG can be viewed as a generalized
least squares methods\cite{DPGOverview}, we might expect it to struggle with
this problem as well. The problem domain is detailed in
Figure~\ref{fig:stokesCylinderDomain} with inlet and outlet velocity profiles
\[
\bfu_{in}=\bfu_{out}=\vecttwo{(1-y)(1+y)}{0}\,,
\]
and zero flow on the cylinder and at the top and bottom walls. We use $\mu=$
with both Stokes problems.

We measure mass loss much more directly in these two Stokes problems. Because
fluid enters and leaves the domain only through the inlet and outlet
boundaries, we should be able to integrate the mass flux over any
cross-section of the mesh and get the same value. Unfortunately, it's not
mathematically kosher to take line integrals of our field variables which only
live in $L^2$. We can, however, integrate the flux over element boundaries.
This carries the unfortunate limitation that we can only measure mass loss
where there is a clear vertical mesh line. We therefore pick integration lines
from the initial coarse mesh and measure the mass flux after each adaptive refinement
step. The percent mass loss is thus
\[
\%m_{loss}=\frac{\int_{\Gamma_{in}}\bfu\cdot\bfn_{in}d\ell-\int_S\bfu\cdot\bfn_Sd\ell}
{\int_{\Gamma_{in}}\bfu\cdot\bfn_{in}d\ell}\times100.
\]

\begin{figure}[p]
\centering
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=2.5cm,y=2.5cm]
\clip(-1.97,-1.48) rectangle (3.39,1.55);
\draw(0,0) circle (1.5cm);
\draw (-1,-1)-- (3,-1);
\draw (3,1)-- (3,-1);
\draw (-1,1)-- (3,1);
\draw (-1,1)-- (-1,-1);
\draw (0,0)-- (.3,0.5196);
\draw (0.12,0.29) node[anchor=north west] {r};
\draw (-1.3,-1.0) node[anchor=north west] {-1.0};
\draw (-1.3,1.15) node[anchor=north west] {1.0};
\draw (-1.3,0.0) node[anchor=north west] {$u_{in}$};
\draw (0.6,0.0) node[anchor=north west] {$u_{cyl}$};
\draw (3.0,-1.0) node[anchor=north west] {3.0};
\draw (3.0,0.0) node[anchor=north west] {$u_{out}$};
\draw (0.9,-1.03) node[anchor=north west] {$u_{wall}$};
\draw (0.9,1.17) node[anchor=north west] {$u_{wall}$};
\end{tikzpicture}
\caption{Stokes cylinder domain}
\label{fig:stokesCylinderDomain}
\end{figure}
